---
title: "Getting started with BalExMed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with BalExMed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

BalExMed fits a mediation model in which a composition, such as the relative
abundances of gut microbial taxa, is the exposure. Existing compositional
mediation methods treat the microbiome as the mediator; here it is upstream of
both the mediator and the outcome.

## Model

The composition $x_i$ enters through a balance, the normalized log-ratio between
the geometric means of two groups of parts,
$$
B(z, x_i) = \sqrt{\frac{a_+ a_-}{a_+ + a_-}}
  \left(\frac{1}{a_+}\sum_{z_j = 1} \log x_{ij} - \frac{1}{a_-}\sum_{z_j = -1} \log x_{ij}\right).
$$
The vector $z \in \{-1, 0, 1\}^d$ is unknown. The mediator $m_i$ and the outcome
$y_i$ follow
$$
m_i = \alpha B(z, x_i) + \psi_m^\top c_i + e_i, \qquad
y_i = \gamma B(z, x_i) + \beta m_i + \psi_y^\top c_i + \varepsilon_i .
$$
For a unit increase in the balance, the direct effect is $\gamma$ and the
indirect effect is $\alpha\beta$. These effects have a causal interpretation
only under the usual mediation assumptions, including no unmeasured confounding
of the exposure, mediator and outcome given $c_i$.

## Simulated example

```{r}
library(BalExMed)
set.seed(2024)
sim <- simulate_balexmed(n = 150, d = 30, alpha = 1/3, beta = 2, gamma = 1/3)
sim$effects
head(sim$z, 8)
```

The data contain no zeros because `simulate_balexmed()` replaces them with a
pseudocount. Real data need the same treatment before fitting.

```{r}
fit <- balexmed(sim$X, sim$m, sim$y, covariates = sim$covariates,
                eta = 1/5, n_iter = 2000, burn_in = 1000,
                n_chains = 2, z_init = "pca", seed = 1)
summary(fit)
```

These chains are short to keep the vignette fast. For an analysis, run several
chains long enough that the split-Rhat values are close to 1 and the inclusion
probabilities agree across chains.

## Selected parts

```{r}
head(taxa_table(fit), 10)
```

`signed` is the posterior probability that a part is in the numerator minus the
probability that it is in the denominator. By construction $\gamma > 0$, so the
numerator holds the parts whose higher relative abundance goes with a higher
outcome through the direct path.

```{r}
z_hat <- selected_z(fit)
table(estimated = z_hat, true = sim$z)
cor(balance(sim$X, z_hat), sim$balance)
```

## Diagnostics

```{r}
plot(fit)
plot(fit, type = "inclusion", top = 12)
```

Posterior draws are available as a data frame or as a `coda` object for further
diagnostics.

```{r}
draws <- effect_draws(fit)
quantile(draws$indirect / draws$total, c(0.025, 0.5, 0.975))
coda::gelman.diag(as_mcmc_list(fit, c("gamma", "indirect")))
```
